Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-02 LICS21/SPCF.tex
Example VI.2 (Labelling samples by terms): typos

\subsection{Sampling semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\Sigma_\entrosp$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}.}

>>>>

\subsection{Sampling semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\Sigma_\entrosp$} and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$.
The maps $\pi_h:\entrosp \to I, \; \pi_t:\entrosp \to \entrosp$ popping the first element are then measurable.
\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and call elements of $\entrosp$ \emph{traces}.}

### 2021-01-02 LICS21/confluent.tex
Example VI.2 (Labelling samples by terms): typos

The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\).

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence. As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics, there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
\qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. This set is unmeasurable, therefore the termination probability would not even be well-defined. Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
\begin{align*}
&\{s \in I^{L_s(A[\tsample])} \mid \\
&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
\end{align*}
which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
\qed
\end{example}

Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed. For example, consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$. If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.

>>>>

\changed[lo]{We are thus lead to consider the reduction relation on skeletons (and positions in a skeleton), which can be extended from the definitions on terms in the obvious way,} 
%The positions in a skeleton and the reduction relation on skeletons can be extended from the definitions on terms in the obvious way, 
with $\tif{\skeletonPlaceholder < 0}{A}{B}$ reducing nondeterministically to both $A$ and $B$, and $\tsample$ reducing to $\skeletonPlaceholder$.
For example, we have 
\(
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \tsample
\to
(\lambda x . \tif{x < 0}{x}{\skeletonPlaceholder}) \, \skeletonPlaceholder
\to
\tif{\skeletonPlaceholder < 0}{\skeletonPlaceholder}{\skeletonPlaceholder}
\to
\skeletonPlaceholder
\).

Given a closed term $M$, let $L_0(M)$ be the set of pairs, the first element of which is a $\to$-reduction sequence of skeletons starting at $Sk(M)$, and the second of which is a position in the final skeleton of the reduction sequence. As with the traces from $I^{\mathbb N}$ used to pre-select samples to use in the usual linear sampling semantics,\lo{I assume by linear sampling semantics, you mean the standard sampling semantics. If so, I would drop linear.} there will be (with one more caveat introduced later) a sample from $I$ pre-selected for each element of $L_0(M)$, which will then be used if a $\tsample$ reduction is ever performed at that position.

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
\qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
\lo{We have just defined $L_0(M)$, but what is $L_s(\cdot)$ (subscript $s$)? I assume it is the version of $L_0(M)$ that labels samples by term.}

\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

\lo{The $\mathsf{if}_1$ in the line above (and in the set comprehension below) should be $\mathsf{if}_1 ; \underline{-}_1$.}

This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. 
This set is unmeasurable, therefore the termination probability would not even be well-defined. 
Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. 
For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
\begin{align*}
&\{s \in I^{L_s(A[\tsample])} \mid \\
&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
\end{align*}
which is measurable. Skeletal reduction sequences still contain just as much information as reduction sequences of terms, however, on which reductions took place.
\qed
\end{example}

Reduction sequences are used rather than reachable skeletons because if the same skeleton is reached twice, different samples may be needed.

\begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$. If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
\end{example}

### 2021-01-02 LICS21/ranking.tex
Example VI.2 (Labelling samples by terms): typos

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
\fi
Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$. The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.

>>>>

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
\fi
Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$.\lo{TODO: reduce the figure to avoid overflow.} The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.
